Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection Fourier transform Mass Spectrometry Experiments

### 2022-04-07 doc/peakcharacterization_manuscript.Rmd
p-value comparisons among different intensities
closes #3

@@

`r figure_count$label_text("nap_intensity")` compares the peak-peak isotopic natural abundance probability and height log-ratio differences (Equation X in Methods) generated using heights from Xcalibur and from our scan-level peak characterization.
:::

@@

::: {custom-style="MDPI_2.2_heading2"}
### Changes in Variance Across a Large Dataset
:::

::: {custom-style="MDPI_3.1_text"}
For large datasets, we expect that more correct peak intensities will result in better agreement between sample normalized peak intensities within a sample class.
If incorrect peak intensities are being reported, then we would expect between class to within class variances to be very random, resulting in F-statistics close to 1.
:::

@@

```{r increment_frequency}
equation_count$increment("mzfrequency")
```

@@

::: {custom-style="MDPI_3.9_equation"}
`r format_equation('frequency = intercept + x* \\frac{1}{\\sqrt{mz}} + y * \\frac{1}{\\sqrt[3]{mz}}', 'mzfrequency')`
:::

::: {custom-style="MDPI_3.9_equation"}
$$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
:::

@@

::: {custom-style="MDPI_3.9_equation"}
$$mz = intercept + x* \frac{1}{frequency} + y*\frac{1}{frequency^2} + z*\frac{1}{frequency^3}$$
:::

@@

::: {custom-style="MDPI_3.9_equation"}
$$\ln{intensity} = intercept + a*position+b*position^2$$
:::

@@

::: {custom-style="MDPI_3.1_text"}
Our previously described SMIRFE algorithm [@mitchellSmallMoleculeIsotope2019] was used to assign molecular formulas to scan-centric characterized peaks in an untargeted manner. 
For the lipid samples, an initial EMF database was generated using an M/Z limit of 1605 M/Z, and maximum numbers for each element were set to C: 130, N: 7, O: 28, P: 3, H: 230. 
Assigned formulas were allowed to have K, Na, H, and NH4 adducts (only positive mode samples were assigned). 
Assigned molecular formulas were then classified into one or more lipid categories using our lipid classifier tool [@mitchellDerivingLipidClassification2020]. 
Lipid classes that were previously observed to be overclassified by our models (neutral and acidic glycosphingolipids) were excluded.
For the ECF derivatized amino-acid samples, the EMF database was generated using an m/z limit of 1005 m/z, and maximum numbers for each element were set to C: 100, N: 7, O: 40, H: 230, P: 3, S: 3.
We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them given its requirements for scan-level peak characterization data.
:::

@@

::: {custom-style="MDPI_3.9_equation"}
$$log(NAP_{P1} / NAP_{P2}) - log(Int_{P1} / Int_{P2}) \approx 0$$
:::

SMIRFE assignments include the NAP for each of the isotopologue molecular formulas (IMFs) in the particular labelled and adducted EMF.
For each peak pair in the peak group for a particular EMF, we calculate the log-ratio for the NAPs and the log-ratio for the intensities, and their absolute differences.
This calculation is done for the final characterized intensity, as well as the intensities at the scan level, and for both the *raw* intensity and *corrected* intensity, and for matched Xcalibur peak intensities, and matched MSnbase peak intensities if there were two or more matching peaks in the labeled, adducted EMF.

>>>>

```{r increment_equations}
equation_count$increment("mzfrequency")
equation_count$increment("frequencymz")
equation_count$increment("modelintensity")
equation_count$increment("napintensity")
```

@@

`r figure_count$label_text("nap_intensity")` compares the peak-peak isotopic natural abundance probability and height log-ratio differences (Equation `r equation_count$label_text("napintensity")` in Methods) generated using heights from Xcalibur and from our scan-centric peak characterization.
:::

@@

::: {custom-style="MDPI_2.2_heading2"}
### Changes in P-Values On a Large Dataset
:::

```{r increment_p_value}
figure_count$increment("p_value_comparison")
table_count$increment("p_value_comparison")
```

::: {custom-style="MDPI_3.1_text"}
For large datasets, we expect that more correct peak intensities will result in better agreement between sample normalized peak intensities within a sample class.
One way to test this is to compare p-values generated from the different intensities.
`r figure_count$label_text("p_value_comparison")` compares the -1 * Log10(p-values) from XX or XX isotope resolved molecular formulas (IMFs) assigned by SMIRFE for the scan-centric peaks and then peaks matched and intensities extracted from the other methods.
This shows that the p-values generated by the raw scan-centric intensities are generally the same if we use the intensities corrected for missing scans (on the line), and the p-values using raw scan-centric are better than either the XCalibur or MSnbase generated peak intensities (below the line due to being worse in XCalibur and MSnbase).
We can test whether these differences are random or not using a binomial test of the actual difference in log-p-values.
The binomial test p-values of the differences are reported in `r table_count$label_text("p_value_comparison")`.
:::

::: {custom-style="MDPI_5.2_figure"}
```{r p_value_comparison, dn_id = figure_count, fig.width = 4, fig.height = 9}
tar_load(lung_compare)

lung_compare$id.test = factor(lung_compare$id.test, c("corrected", "xcalibur", "msnbase"))

lung_compare %>%
  ggplot(aes(x = log_p.ref,
             y = log_p.test)) +
  geom_point() +
  geom_abline(slope = 1, color = "red") +
  facet_wrap(~ id.test, ncol = 1) +
  labs(x = "-1 * Log10(P-Value) Raw Scan-Centric",
       y = "-1 * Log10(P-Value) Other")
```
:::

::: {custom-style="MDPI_5.1_figure_caption"}
`r figure_count$label_text("p_value_comparison")`.
Log-p-values generated by comparing non-cancer and cancer sample IMFs using peak intensities from different methods.
:::

::: {custom-style="MDPI_4.1_table_caption"}
`r table_count$label_text("p_value_comparison")`.
Binomial test results on log-p-value differences to p-values generated by raw scan-centric IMF intensities.
P-values were adjusted using the Bonferroni method.
:::

::: {custom-style="MDPI_4.2_table_body"}
```{r p_value_comparison_table}
tar_load(lung_binomial)
binomial_order = data.frame("comparison" = c("corrected", "xcalibur", "msnbase"))
lung_binomial = dplyr::left_join(binomial_order,
                                 lung_binomial,
                                 by = "comparison")
lung_binomial %>%
  dplyr::transmute(Comparison = comparison,
                `N-Positive` = n_positive,
                `N-Negative` = n_negative,
                `P-Value` = sprintf("%.02e", p.value),
                `Adj-P-Value` = sprintf("%.02e", p.adjust)) %>%
  flextable() %>%
  align(j = c(4, 5), align = "right", part = "all") %>%
  autofit()
```
:::

@@

@@

::: {custom-style="MDPI_3.9_equation"}
`r format_equation('frequency = a + x* \\frac{1}{\\sqrt{mz}} + y * \\frac{1}{\\sqrt[3]{mz}}', 'mzfrequency')`
:::

@@

::: {custom-style="MDPI_3.9_equation"}
`r format_equation('mz = a + x * \\frac{1}{frequency} + y * \\frac{1}{frequency^2} + z *\\frac{1}{frequency^3}', 'frequencymz')`
:::

@@

::: {custom-style="MDPI_3.9_equation"}
`r format_equation('\\ln{intensity} = a + x * position + y * position^2', 'modelintensity')`
:::

@@

::: {custom-style="MDPI_3.1_text"}
Our previously described SMIRFE algorithm [@mitchellSmallMoleculeIsotope2019] was used to assign molecular formulas to scan-centric characterized peaks in an untargeted manner. 
For the lipid samples, an initial EMF database was generated using an m/z limit of 1605 m/z, and maximum numbers for each element were set to C: 130, N: 7, O: 28, P: 3, H: 230. 
Assigned formulas were allowed to have K, Na, H, and NH4 adducts (only positive mode samples were assigned). 
Assigned molecular formulas were then classified into one or more lipid categories using our lipid classifier tool [@mitchellDerivingLipidClassification2020]. 
Lipid classes that were previously observed to be overclassified by our models (neutral and acidic glycosphingolipids) were excluded.
For the ECF derivatized amino-acid samples, the EMF database was generated using an m/z limit of 1005 m/z, and maximum numbers for each element were set to C: 100, N: 7, O: 40, H: 230, P: 3, S: 3.
We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them given its requirements for scan-level peak characterization data.
:::

@@

::: {custom-style="MDPI_3.9_equation"}
`r format_equation('\\ln(NAP_{P1} / NAP_{P2}) - \\ln(Int_{P1} / Int_{P2}) \\approx 0', 'napintensity')`
:::

::: {custom-style="MDPI_3.1_text"}
SMIRFE assignments include the NAP for each of the isotopologue molecular formulas (IMFs) in the particular labelled and adducted EMF.
For each peak pair in the peak group for a particular EMF, we calculate the log-ratio for the NAPs and the log-ratio for the intensities, and their absolute differences.
This calculation is done for the final characterized intensity, as well as the intensities at the scan level, and for both the *raw* intensity and *corrected* intensity, and for matched Xcalibur peak intensities, and matched MSnbase peak intensities if there were two or more matching peaks in the labeled, adducted EMF.
:::
